Improving Software Maintainability through Automated Refactoring of Code Clones

### 2019-08-24 icse/paper.tex
Cite

\section{Introduction}
Duplicate fragments in source code (also named ``code clones'') are often seen as one of the most harmful types of technical debt \cite{fowler2018refactoring}. Duplicate fragments create implicit dependencies that make the code harder to maintain as the resolution of erroneous behaviour in one location may have to be applied to all the cloned code as well \cite{ostberg2014automatically}. Apart from that, code clones can contribute up to 25\% of the code size~\cite{bruntink2005use}.

@@

In this study, we define a technique to detect clones such that they can be automatically refactored. We list a set of criteria that can obtruct refactoring opportunities. We evaluate the relation and location of clones to determine which refactoring techniques are best suited. We introduce a tool that automatically refactors clones by extracting new methods out of duplicated code.

@@

\section{Background}
Our study lies intersection of code clone and refactoring research. We use two concepts to argue about code clones \cite{roy2007survey}:
\\ \textbf{Clone instance}: A single cloned fragment.
\\ \textbf{Clone class}: A set of similar clone instances.

The most common technique for refactoring clones is ``Extract Method'' \cite{fowler2018refactoring}, which can be applied on code inside the body of a method. This technique moves functionality in method bodies to a new method. This technique extracts the contents of a single clone instance to a new method and replace all clone instances in a clone class by a call to this method.

\section{Defining refactorable clones}
In literature, several clone type definitions have been used to argue about duplication in source code \cite{roy2007survey}. We discuss how we can define clones such that they can be automatically refactored without side effects on the source code.

\subsection{Ensuring Equality}\label{sec:t1r}
Most modern clone detection tools detect duplication by comparing the code textually together with the omission of certain tokens \cite{roy2009comparison, svajlenko2014evaluating}. Clones detected by such means may not always be suitable for refactoring, because textual comparison fails to take into account the context of certain symbols in the code. Information that gets lost in textual comparison is the referenced declaration for type, variable and method references. Equally named type, variable and method references may refer to different declarations with a different implementation. Such clones can be harder to refactor, if beneficial at all.

@@

\subsection{Allowing variability in a controlled set of expressions} \label{sec:t2r}
Often, duplication fragments in source code do not match exactly \cite{kodhai2013method}. When developers duplicate fragments of code, they modify the duplicated block to fit its new location and purpose. To detect duplicate fragments with minor variance, we look into what expressions we can allow variability in, while still being refactorable.

We define the following expressions as refactorable when varied:
\begin{itemize}
  \item \textbf{Literals}: Only if all varying literals in a clone class have the same type.
  \item \textbf{Variables}: Only if all varying variables in a clone class have the same type.
  \item \textbf{Method references}: Only if the return value of referenced methods match (or are not used).
\end{itemize}
Often when allowing such variance, trade-offs come into play \cite{krishnan2013refactoring, krishnan2014unification}. For instance, variance in literals may require the introduction of a parameter to an extracted method if the ``Extract Method'' refactoring method is used, increasing the required effort to comprehend the code.

\subsection{Gapped clones} \label{sec:t3r}
Sometimes, when fragments are duplicated, a statement is inserted or changed severely for the code to fit its new context \cite{roy2007survey}. When dealing with such a situation, there are several opportunities to refactor so-called ``gapped clones'' \cite{ueda2002detection, zhao2018automatic}. ``Gapped clones'' are two clone instances separated by a ``gap'' of non-cloned statement(s). We define the following methods to refactor such clones:
\begin{itemize}
  \item We wrap the difference in statements in a conditionally executed block, one path for each different (group of) statement(s).
  \item We use a lambda function to pass the difference in statements from each location of the clone \cite{tsantalis2017clone}.
\end{itemize}
For both refactoring techniques a trade-off is at play, as these solutions increase the complexity and volume of the source code in favor of removing a clone.

\section{The CloneRefactor tool}
To detect such clones that can be refactored, we surveyed a set of modern clone detection tools and techniques \cite{svajlenko2014evaluating, sheneamer2016survey, gautam2016various, roy2009comparison}. We created a set of control projects\footnote{Control projects for assessing clone detection tools: \url{https://github.com/SimonBaars/CloneRefactor/tree/master/src/test/resources}} to determine the suitability of these tools to detect refactorable clones, either through configuration or postprocessing of their output.

None of the surveyed tools \cite{ragkhitwetsagul2019siamese, roy2008nicad, semura2017ccfindersw, sajnani2016sourcerercc, saini2018oreo, jiang2007deckard, kamalpriya2017enhancing, mazinanian2016jdeodorant} were suitable with our findings about refactorable clones, because of which we decided to implement our own tool: CloneRefactor\footnote{The source code of CloneRefactor is available on GitHub: \url{https://github.com/SimonBaars/CloneRefactor}}. This tool goes through a 3-step process to automatically refactor clones as shown in Fig.~\ref{fig:clonerefactorflow}. In the following sections, we explain these steps.
\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{img/flow}
  \caption{Simple flow diagram of CloneRefactor.}
  \label{fig:clonerefactorflow}
\end{figure}

@@

\subsubsection{Relation}\label{sec:relation}
Clone instances in a clone class may have a relation with each other through inheritance. This relation has a big impact on how the clone should be refactored \cite{fontana2015duplicated}. We define the following categories to map the relation between clone instances in a clone class. These categories do not map external classes (classes outside the project, e.g. belonging to a library) unless explicitly stated:
\begin{itemize}
    \item \textbf{Common Class}: All clone instances are in the same class.
    \begin{itemize}
        \item \textbf{Same Method}: All clone instances are in the same method.
        \item \textbf{Same Class}: All clone instances are in the same class.
    \end{itemize}
    \item \textbf{Common Hierarchy}: All clone instances are in the same inheritance hierarchy.
    \begin{itemize}
        \item \textbf{Superclass}: Clone instances reside in a class or its parent class.
        \item \textbf{Sibling Class}: All clone instances reside in classes with the same parent class.
        \item \textbf{Ancestor Class}: All clone instances reside in a class, or any of its recursive parents.
        \item \textbf{First Cousin}: All clone instances reside in classes with the same grandparent class.
        \item \textbf{Same Hierarchy}: All clone instances are part of the same inheritance hierarchy.
    \end{itemize}
    \item \textbf{Common Interface}: All clone instances are in classes that have the same interface.
    \begin{itemize}
        \item \textbf{Same Direct Interface}: All clone instances are in a class that have the same interface.
        \item \textbf{Same Class}: All clone instances are in an inheritance hierarchy that contains the same interface.
    \end{itemize}
    \item \textbf{Unrelated}: All clone instances are in classes that have the same interface.
    \begin{itemize}
        \item \textbf{No Direct Superclass}: All clone instances are in classes that have the Object class as parent.
        \item \textbf{No Indirect Superclass}: All clone instances are in a hierarchy that contains a class that has the Object class as parent.
        \item \textbf{External Superclass}: All clone instances are in classes the same external class as parent.
        \item \textbf{Indirect External Ancestor}: All clone instances are in a hierarchy that contains a class that has an external class as parent.
    \end{itemize}
\end{itemize}

@@

\subsubsection{AST Transformation}
CloneRefactor uses JavaParser \cite{tomassetti2017javaparser}: an AST-parsing library that allows to modify the AST and write it back to source code. To refactor clones, CloneRefactor creates a new method declaration and moves all statements from a clone instance in the clone class to the new method. This method is placed according to the relation between the clone instances (see Sec.~\ref{sec:relation}). CloneRefactor analyzes the source code of the extracted method and populates it with the following properties:
\begin{itemize}
  \item \textbf{Parameters}: For each variable used that is not accessible from the scope of the extracted method.
  \item \textbf{A return value}: If the method modifies or declares local data that is needed outside of its scope, or if the cloned fragments already returned data.
  \item \textbf{Thrown exception}: If the method throws an uncaught exception that is not a \texttt{RuntimeException}.
\end{itemize}
CloneRefactor then removes all cloned code and replaces it with a method call to the newly created method. In case of a return value, CloneRefactor either assigns the call result, declares it or returns it accordingly.

@@

\subsubsection{Impact on maintainability metrics} \label{sec:metrics}
CloneRefactor measures the impact on maintainability metrics of the refactored source code for each clone class that is refactored. These metrics are derived from Heitlager et al. \cite{heitlager2007practical}. This paper defines a set of metrics to measure the maintainability of a system. For each of these metrics, risk profiles are proposed to determine the maintainability impact on the system of a whole.

To determine whether the maintainability improves when refactoring a single clone, we need to measure the impact of small-grained changes. Because of that, we measure only a subset of the metrics \cite{heitlager2007practical} and focus on the absolute metric changes (instead of the risk profiles). The subset of metrics we decided to focus on are all metrics that are measured on method level (as the other metrics show a lesser impact on the maintainability for these small changes). These metrics are:
\begin{itemize}
\item \textbf{Duplication}: In Heitlager et al. \cite{heitlager2007practical} this metric is measured by taking the amount of duplicated lines. We decided to use the amount of duplicated tokens part of a clone class instead, to have a stronger reflection of the impact of the refactoring by measuring a more fine-grained system property.
\item \textbf{Volume}: The more code a system has, the more code has to be maintained. The paper \cite{heitlager2007practical} measures volume as lines of code. As with duplication, we use the number of tokens instead.
\item \textbf{Complexity}: Heitlager et al. use MCCabe complexity \cite{mccabe1976complexity} to calculate their complexity metric. The MCCabe complexity is a quantitative measure of the number of linearly independent paths through a method.
\item \textbf{Method Interface Size}: The number of parameters that a method has. If a method has many parameters, the code may become harder to understand and it is an indication that data is not encapsulated adequately in objects \cite{fowler2018refactoring}.
\end{itemize}

@@

\subsection{Corpus}
We ran all our experiments using CloneRefactor on a corpus of open source Java projects. This corpus is derived from the corpus of a study that uses machine learning to determine the suitability of Java projects on GitHub for data analysis purposes \cite{githubCorpus2013}.

@@

\subsection{Thresholds}
Most clone detection tools can be configured using thresholds. These thresholds indicate the minimum number of lines, tokens and/or statements that must be spanned for duplicate fragments to be considered clones. Often, such thresholds are intuitively chosen \cite{li2006cp, roy2009mutation} or based on a quartile distribution of empirical data \cite{alves2010deriving}. Using the maintainability score we can find support for which thresholds should be chosen to increase the chance to find clones that improve maintainability when refactored.

>>>>

\section{Introduction}
Duplicate fragments in source code (also named ``code clones'') are often seen as one of the most harmful types of technical debt~\cite{fowler2018refactoring}. Duplicate fragments create implicit dependencies that make the code harder to maintain as the resolution of erroneous behaviour in one location may have to be applied to all the cloned code as well~\cite{ostberg2014automatically}. Apart from that, code clones can contribute up to 25\% of the code size~\cite{bruntink2005use}.

@@

In this study, we define a technique to detect clones such that they can be automatically refactored. We list a set of criteria that can obstruct refactoring opportunities. We evaluate the relation and location of clones to determine which refactoring techniques are best suited. We introduce a tool that automatically refactors clones by extracting new methods out of duplicated code.

@@

\section{Background}
Our study lies intersection of code clone and refactoring research. We use two concepts to argue about code clones~\cite{roy2007survey}:
\\ \textbf{Clone instance}: A single cloned fragment.
\\ \textbf{Clone class}: A set of similar clone instances.

The most common technique for refactoring clones is ``Extract Method''~\cite{fowler2018refactoring}, which can be applied on code inside the body of a method. This technique moves functionality in method bodies to a new method. This technique extracts the contents of a single clone instance to a new method and replace all clone instances in a clone class by a call to this method.

\section{Defining refactorable clones}
In literature, several clone type definitions have been used to argue about duplication in source code~\cite{roy2007survey}. We discuss how we can define clones such that they can be automatically refactored without side effects on the source code.

\subsection{Ensuring Equality}\label{sec:t1r}
Most modern clone detection tools detect duplication by comparing the code textually together with the omission of certain tokens~\cite{roy2009comparison, svajlenko2014evaluating}. Clones detected by such means may not always be suitable for refactoring, because textual comparison fails to take into account the context of certain symbols in the code. Information that gets lost in textual comparison is the referenced declaration for type, variable and method references. Equally named type, variable and method references may refer to different declarations with a different implementation. Such clones can be harder to refactor, if beneficial at all.

@@

\subsection{Allowing variability in a controlled set of expressions} \label{sec:t2r}
Often, duplication fragments in source code do not match exactly~\cite{kodhai2013method}. When developers duplicate fragments of code, they modify the duplicated block to fit its new location and purpose. To detect duplicate fragments with minor variance, we look into what expressions we can allow variability in, while still being refactorable.

We define the following expressions as refactorable when varied:
\begin{itemize}
  \item \textbf{Literals}: Only if all varying literals in a clone class have the same type.
  \item \textbf{Variables}: Only if all varying variables in a clone class have the same type.
  \item \textbf{Method references}: Only if the return value of referenced methods match (or are not used).
\end{itemize}
Often when allowing such variance, trade-offs come into play~\cite{krishnan2013refactoring, krishnan2014unification}. For instance, variance in literals may require the introduction of a parameter to an extracted method if the ``Extract Method'' refactoring method is used, increasing the required effort to comprehend the code.

\subsection{Gapped clones} \label{sec:t3r}
Sometimes, when fragments are duplicated, a statement is inserted or changed severely for the code to fit its new context~\cite{roy2007survey}. When dealing with such a situation, there are several opportunities to refactor so-called ``gapped clones''~\cite{ueda2002detection, zhao2018automatic}. ``Gapped clones'' are two clone instances separated by a ``gap'' of non-cloned statement(s). We define the following methods to refactor such clones:
\begin{itemize}
  \item We wrap the difference in statements in a conditionally executed block, one path for each different (group of) statement(s).
  \item We use a lambda function to pass the difference in statements from each location of the clone~\cite{tsantalis2017clone}.
\end{itemize}
For both refactoring techniques a trade-off is at play, as these solutions increase the complexity and volume of the source code in favor of removing a clone.

\section{The CloneRefactor tool}
To detect such clones that can be refactored, we surveyed a set of modern clone detection tools and techniques~\cite{svajlenko2014evaluating, sheneamer2016survey, gautam2016various, roy2009comparison}. We created a set of control projects\footnote{Control projects for assessing clone detection tools: \url{https://github.com/SimonBaars/CloneRefactor/tree/master/src/test/resources}} to determine the suitability of these tools to detect refactorable clones, either through configuration or postprocessing of their output.

None of the surveyed tools~\cite{ragkhitwetsagul2019siamese, roy2008nicad, semura2017ccfindersw, sajnani2016sourcerercc, saini2018oreo, jiang2007deckard, kamalpriya2017enhancing, mazinanian2016jdeodorant} were suitable with our findings about refactorable clones, because of which we decided to implement our own tool: CloneRefactor\footnote{The source code of CloneRefactor is available on GitHub: \url{https://github.com/SimonBaars/CloneRefactor}}. This tool goes through a 3-step process to automatically refactor clones as shown in Fig.~\ref{fig:clonerefactorflow}. In the following sections, we explain these steps.
\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{img/flow}
  \caption{Simple flow diagram of CloneRefactor.}
  \label{fig:clonerefactorflow}
\end{figure}

@@

\subsubsection{Relation}\label{sec:relation}
Clone instances in a clone class may have a relation with each other through inheritance. This relation has a big impact on how the clone should be refactored~\cite{fontana2015duplicated}. We define the following categories to map the relation between clone instances in a clone class. These categories do not map external classes (classes outside the project, e.g. belonging to a library) unless explicitly stated:
\begin{itemize}
    \item \textbf{Common Class}: All clone instances are in the same class.
    \begin{itemize}
        \item \textbf{Same Method}: All clone instances are in the same method.
        \item \textbf{Same Class}: All clone instances are in the same class.
    \end{itemize}
    \item \textbf{Common Hierarchy}: All clone instances are in the same inheritance hierarchy.
    \begin{itemize}
        \item \textbf{Superclass}: Clone instances reside in a class or its parent class.
        \item \textbf{Sibling Class}: All clone instances reside in classes with the same parent class.
        \item \textbf{Ancestor Class}: All clone instances reside in a class, or any of its recursive parents.
        \item \textbf{First Cousin}: All clone instances reside in classes with the same grandparent class.
        \item \textbf{Same Hierarchy}: All clone instances are part of the same inheritance hierarchy.
    \end{itemize}
    \item \textbf{Common Interface}: All clone instances are in classes that have the same interface.
    \begin{itemize}
        \item \textbf{Same Direct Interface}: All clone instances are in a class that have the same interface.
        \item \textbf{Same Class}: All clone instances are in an inheritance hierarchy that contains the same interface.
    \end{itemize}
    \item \textbf{Unrelated}: All clone instances are in classes that have the same interface.
    \begin{itemize}
        \item \textbf{No Direct Superclass}: All clone instances are in classes that have the Object class as parent.
        \item \textbf{No Indirect Superclass}: All clone instances are in a hierarchy that contains a class that has the Object class as parent.
        \item \textbf{External Superclass}: All clone instances are in classes the same external class as parent.
        \item \textbf{Indirect External Ancestor}: All clone instances are in a hierarchy that contains a class that has an external class as parent.
    \end{itemize}
\end{itemize}

@@

\subsubsection{AST Transformation}
CloneRefactor uses JavaParser~\cite{tomassetti2017javaparser}: an AST-parsing library that allows to modify the AST and write it back to source code. To refactor clones, CloneRefactor creates a new method declaration and moves all statements from a clone instance in the clone class to the new method. This method is placed according to the relation between the clone instances (see Sec.~\ref{sec:relation}). CloneRefactor analyzes the source code of the extracted method and populates it with the following properties:
\begin{itemize}
  \item \textbf{Parameters}: For each variable used that is not accessible from the scope of the extracted method.
  \item \textbf{A return value}: If the method modifies or declares local data that is needed outside of its scope, or if the cloned fragments already returned data.
  \item \textbf{Thrown exception}: If the method throws an uncaught exception that is not a \texttt{RuntimeException}.
\end{itemize}
CloneRefactor then removes all cloned code and replaces it with a method call to the newly created method. In case of a return value, CloneRefactor either assigns the call result, declares it or returns it accordingly.

@@

\subsubsection{Impact on maintainability metrics} \label{sec:metrics}
CloneRefactor measures the impact on maintainability metrics of the refactored source code for each clone class that is refactored. These metrics are derived from Heitlager et al.~\cite{heitlager2007practical}. This paper defines a set of metrics to measure the maintainability of a system. For each of these metrics, risk profiles are proposed to determine the maintainability impact on the system of a whole.

To determine whether the maintainability improves when refactoring a single clone, we need to measure the impact of small-grained changes. Because of that, we measure only a subset of the metrics~\cite{heitlager2007practical} and focus on the absolute metric changes (instead of the risk profiles). The subset of metrics we decided to focus on are all metrics that are measured on method level (as the other metrics show a lesser impact on the maintainability for these small changes). These metrics are:
\begin{itemize}
\item \textbf{Duplication}: In Heitlager et al.~\cite{heitlager2007practical} this metric is measured by taking the amount of duplicated lines. We decided to use the amount of duplicated tokens part of a clone class instead, to have a stronger reflection of the impact of the refactoring by measuring a more fine-grained system property.
\item \textbf{Volume}: The more code a system has, the more code has to be maintained. The paper~\cite{heitlager2007practical} measures volume as lines of code. As with duplication, we use the number of tokens instead.
\item \textbf{Complexity}: Heitlager et al.~\cite{heitlager2007practical} use MCCabe complexity~\cite{mccabe1976complexity} to calculate their complexity metric. The MCCabe complexity is a quantitative measure of the number of linearly independent paths through a method.
\item \textbf{Method Interface Size}: The number of parameters that a method has. If a method has many parameters, the code may become harder to understand and it is an indication that data is not encapsulated adequately in objects~\cite{fowler2018refactoring}.
\end{itemize}

@@

\subsection{Corpus}
We ran all our experiments using CloneRefactor on a corpus of open source Java projects. This corpus is derived from the corpus of a study that uses machine learning to determine the suitability of Java projects on GitHub for data analysis purposes~\cite{githubCorpus2013}.

@@

\subsection{Thresholds}
Most clone detection tools can be configured using thresholds. These thresholds indicate the minimum number of lines, tokens and/or statements that must be spanned for duplicate fragments to be considered clones. Often, such thresholds are intuitively chosen~\cite{li2006cp, roy2009mutation} or based on a quartile distribution of empirical data~\cite{alves2010deriving}. Using the maintainability score we can find support for which thresholds should be chosen to increase the chance to find clones that improve maintainability when refactored.
